---
title: "Project"
output:
  html_document:
    df_print: paged
    toc: yes
    toc_depth: '2'
  pdf_document:
    toc: yes
    toc_depth: 2
---

<style>
body {
text-align: justify}
</style>

```{css, echo=FALSE}
.solution {
background-color: #CCDDFF;
}
```

\def\bx{\mathbf{x}}

**Note: have not set seeds; have not split train and test datasets, have not finished data preprocessing**   

# Project Description (est. 1 page, pt. 5)

# Literature Review(est. 1 page, pt.10)

# preprocessing and Summary statistics(est. 1 -2 pages, pt.10)
- check multicollinearity 
- check outliers (`rs` and `pp`)
- check balance (`cn` and `mu`)

# models that perform well on high-dimensional data
- SVM 
- Random Forest 
- Lasso (?)
- KNN regression

```{r}
brca = read.csv("brca_data_w_subtypes.csv")
# head(brca)
```

```{r}
dim(brca) # 705 rows, 1941 columns
names(brca)[1937:1941] # outcomes 
# 1936 covariates: 860 copy number variations (cn), 249 somatic mutations (mu), 604 gene expressions (rs), and 223 protein levels (pp)
# order: rs 1:604, cn 605:1464, mu 1465:1713, pp 1714:1936

brca = brca[,-1937] # discard `vital.status` 
names(brca)[1937:1940]
```

```{r}
# unique values of responses 
unique(brca$PR.Status)
unique(brca$ER.Status)
unique(brca$HER2.Final.Status)
unique(brca$histological.type)
```

```{r}
# only use Negative and Positive for PR.status, ER.status, and HER2.Final.Status
# PR.status and ER.status are highly correlated
table(brca$PR.Status) 
table(brca$ER.Status)
table(brca$HER2.Final.Status)
table(brca$histological.type)
```


```{r}
# sub is the dataset we will use for modeling 
# all of the null values are removed 
sub = brca[(brca$PR.Status == "Positive" | brca$PR.Status == "Negative") & 
           (brca$ER.Status == "Positive" | brca$ER.Status == "Negative") & 
           (brca$HER2.Final.Status == "Positive" | 
            brca$HER2.Final.Status == "Negative"),]
dim(sub)
```

```{r}
# the input variables have the indices below 
# rs 1:604, cn 605:1464, mu 1465:1713, pp 1714:1936
```

check correlation 
```{r}
rs = sub[1:604] # the subset that only contains rs
corr = round(cor(rs), 2) # correlation matrix 
idx = data.frame(NA, NA, NA)
for (i in 1:nrow(corr)) {
  for (j in 1:nrow(corr)) {
    if (abs(corr[i, j]) > 0.8 & i < j) {
      idx[nrow(idx) + 1,] = c(i, j, corr[i, j])
    }
  }
}
idx = idx[-1,] # stores correlations that are greater than 0.8 -> multicollinearity 
dim(idx)
names(idx) = c("i", "j", "corr")

# remove highly-correlated variables 
rmv = idx[idx$j %in% names(sort(table(idx$j), decreasing = T)[1:61]),]$i 
rmv = unique(rmv)
length(rmv)

rs = rs[,-rmv]
```

```{r}
cn = sub[605:1464]
corr = round(cor(cn), 2)
idx = data.frame(NA, NA, NA)
for (i in 1:nrow(corr)) {
  for (j in 1:nrow(corr)) {
    if (abs(corr[i, j]) > 0.8 & i < j) {
      idx[nrow(idx) + 1,] = c(i, j, corr[i, j])
    }
  }
}
idx = idx[-1,]
dim(idx)
names(idx) = c("i", "j", "corr")

rmv = idx[idx$j %in% names(sort(table(idx$j), decreasing = T)[1:683]),]$i
rmv = unique(rmv)
length(rmv)

cn = cn[,-rmv]
```

```{r}
mu = sub[1465:1713]
corr = round(cor(mu), 2)
idx = data.frame(NA, NA, NA)
for (i in 1:nrow(corr)) {
  for (j in 1:nrow(corr)) {
    if (abs(corr[i, j]) > 0.8 & i < j) {
      idx[nrow(idx) + 1,] = c(i, j, corr[i, j])
    }
  }
}
idx = idx[-1,]
dim(idx)
names(idx) = c("i", "j", "corr")

rmv = idx[idx$j %in% names(sort(table(idx$j), decreasing = T)[1:10]),]$i
rmv = unique(rmv)
length(rmv)
# there is no multicollinearity within mu, so no variable is removed here 
# mu = mu[,-rmv]
```

```{r}
pp = sub[1714:1936]
corr = round(cor(pp), 2)
idx = data.frame(NA, NA, NA)
for (i in 1:nrow(corr)) {
  for (j in 1:nrow(corr)) {
    if (abs(corr[i, j]) > 0.8 & i < j) {
      idx[nrow(idx) + 1,] = c(i, j, corr[i, j])
    }
  }
}
idx = idx[-1,]
dim(idx)
names(idx) = c("i", "j", "corr")

rmv = idx[idx$j %in% names(sort(table(idx$j), decreasing = T)[1:683]),]$i
rmv = unique(rmv)
length(rmv)

pp = pp[,-rmv]
```

# Histological Type(Modeling, SVM and random Forest) (est. 2-3 pages, pt.20)
```{r}
y = as.factor(sub$histological.type)
y = ifelse(y == "infiltrating lobular carcinoma", 1, 0)
sub2 = cbind(rs, cn, mu, pp, y) # cleaned dataset with histoological.type as response 
dim(sub2)
```

```{r}
set.seed(651978735) 
n = dim(sub)[1]
test_size = as.integer(0.25 * n)
test_idx = sample(1:n, test_size) # 25% of the sample size 
test_idx

Xtest = sub2[test_idx, -1]
Xtrain = sub2[-test_idx, -1]
```


```{r}
library(e1071)
svm.fit = svm(y ~., data=sub2, 
              type="C-classification", kernel="linear", scale=F, cost=1)
table("fitted" = svm.fit$fitted, "actual" = y)
```

```{r}
library(ROCR)
roc = prediction(as.numeric(svm.fit$fitted), sub2$y)
performance(roc, measure = "auc")@y.values[[1]]

perf = performance(roc, "tpr", "fpr") 
plot(perf, colorize = T)
```

```{r}
library(randomForest)
rf.fit = randomForest(sub2[1:1098], as.factor(sub2$y), 
                      ntree=500, 
                      mtry=10, 
                      nodesize=10, 
                      samplesize=400, 
                      importance=TRUE)
```

```{r}
pred = predict(rf.fit, sub2[1:1098])
table("fitted" = pred, "actual" = sub2$y)
roc = prediction(as.numeric(pred), sub2$y)
performance(roc, measure = "auc")@y.values[[1]]

perf = performance(roc, "tpr", "fpr") 
plot(perf, colorize = T)
```

see how SVM and Random Forest perform on the raw data `sub` 
```{r}
# test raw data
test = randomForest(sub[1:1936], as.factor(sub2$y), 
                      ntree=500, 
                      mtry=10, 
                      nodesize=10, 
                      samplesize=400, 
                      importance=TRUE)

pred = predict(test, sub[1:1936])
table("fitted" = pred, "actual" = sub2$y)
roc = prediction(as.numeric(pred), sub2$y)
performance(roc, measure = "auc")@y.values[[1]]

perf = performance(roc, "tpr", "fpr") 
plot(perf, colorize = T)
```

```{r}
# test raw data 
test = svm(as.factor(y) ~., data=sub[1:1936], 
              type="C-classification", kernel="linear", scale=F, cost=1)
table("fitted" = test$fitted, "actual" = y)
roc = prediction(as.numeric(svm.fit$fitted), y)
performance(roc, measure = "auc")@y.values[[1]]

perf = performance(roc, "tpr", "fpr") 
plot(perf, colorize = T)
```

# PR Status (to be decided) (est 2-3 pages, pt.20)
```{r}
y = as.factor(sub$PR.Status)
y = ifelse(y == "Positive", 1, 0)
sub3 = cbind(rs, cn, mu, pp, y) # cleaned dataset with PR.status as response 
dim(sub2)
```

```{r}
svm.fit = svm(y ~., data=sub3, 
              type="C-classification", kernel="linear", scale=F, cost=1)
table("fitted" = svm.fit$fitted, "actual" = y)

# library(ROCR)
roc = prediction(as.numeric(svm.fit$fitted), sub3$y)
performance(roc, measure = "auc")@y.values[[1]]

perf = performance(roc, "tpr", "fpr") 
plot(perf, colorize = T)
```

# Variable Selection for All Outcomes (random forest?) (est. 2-3 pages. pt.20)

Using Random Forest, select the most important 50 variables, and make predictions based on these variables. 
```{r}
impt = importance(rf.fit)[order(importance(rf.fit)[,3], decreasing=TRUE),][1:50,]
vars = rownames(impt)
```

```{r}
# sub4 is the cleaned dataset with all four response variables
sub4 = subset(sub, select = vars)
sub4 = cbind(sub4, sub[1937:1940])
sub4$PR.Status = as.factor(sub4$PR.Status)
sub4$histological.type = as.factor(sub4$histological.type)
sub4$ER.Status = as.factor(sub4$ER.Status)
sub4$HER2.Final.Status = as.factor(sub4$HER2.Final.Status)
```

```{r}
X = sub4[1:50] # input variables 
svm.fit = svm(sub4$ER.Status ~., data=X, 
              type="C-classification", kernel="linear", scale=F, cost=1)
table("fitted" = svm.fit$fitted, "actual" = y)

# library(ROCR)
roc = prediction(as.numeric(svm.fit$fitted), sub4$ER.Status)
performance(roc, measure = "auc")@y.values[[1]]

perf = performance(roc, "tpr", "fpr") 
plot(perf, colorize = T)
```

```{r}
rf.fit = randomForest(X, sub4$HER2.Final.Status, 
                      ntree=500, 
                      mtry=7, 
                      nodesize=10, 
                      samplesize=400, 
                      importance=TRUE)

pred = predict(rf.fit, X)
table("fitted" = pred, "actual" = sub4$HER2.Final.Status)
roc = prediction(as.numeric(pred), sub4$HER2.Final.Status)
performance(roc, measure = "auc")@y.values[[1]]

perf = performance(roc, "tpr", "fpr") 
plot(perf, colorize = T)
```


